---
title: "Analysis of Cholesterol Levels in the United States"
output: html_notebook
author: "Juan Carballo"
class: "Data Visualization - Final Project"
---
## Data visualization - Final Project
***
### Source: National Health and Nutrition Examination Survey 
#### https://www.cdc.gov/nchs/index.htm
#### The National Health and Nutrition Examination Survey is a program designed to asses the health and nutritional status of adults and children in the United States. NHANES is a major program of the National Center for Health Statistics (NCHS). NCHS is part of the Centers for Disease Control and Prevention (CDC).

#### The survey examines a sample of 5,000 persons each year, located in counties across the country. NHANES interview includes demographic, socioeconomic, dietary, and health-related questions. Examination component consists of medical, dental, physiological measurements and laboratory tests.
***
#### NHANES 2015–2016 data source was selected since it includes LDL and Triglycerides needed for analysis
### Data source was downloaded from the following link: 
#### https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Laboratory&CycleBeginYear=2015
### Data source documentation, codebook and frequencies
#### Demographic data from surveys:
#### https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DEMO_I.htm
####
#### Cholesterol - Low Density Lipoprotein (LDL) and Triglycerides lab results: 
#### https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/TRIGLY_I.htm
####
#### Cholesterol - High Density Lipoprotein (HDL) lab results: 
#### https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/HDL_I.htm
####
***

### Loading R libraries
```{r}
library(tidyverse) #loads ggplot, dplyr, readr
library(SASxport) #loads data in xport format - SAS
library(plotly) #builds interactive graphs
library(gifski)
library(gganimate)
library(leaflet)
```

### Loading cholesterol - LDL and Triglycerides lab results
```{r}
trigly <- read.xport("../data/TRIGLY_I.XPT")
```

### Loading cholesterol - HDL lab results
```{r}
hdl <- read.xport("../data/HDL_I.XPT")
```

### Loading demographic data
```{r}
demo <- read.xport("../data/DEMO_I.XPT")
```

### Loading map data from usa map
```{r}
usa <- map_data("usa")
```

### Exploratory data analysis
#### Coding heart disease risk based on LDL score
```{r}
trigly <- trigly %>%
  mutate(
    RISK = case_when(
      LBDLDL <= 130 ~ "low",
      LBDLDL >= 190 ~ "high",
      TRUE ~ "medium"
    )
  )
# table with risk values
table(trigly$RISK)
```

#### Getting race and gender demographics about patients with LDL data
```{r}
# joining trigly data with demographic by respondent sequence number
demo_trigly <- trigly %>%
  inner_join(demo) %>%
  mutate(
    RIDRETH1 = case_when(
      RIDRETH1 == 1 ~ "Mexican American",
      RIDRETH1 == 2 ~ "Other Hispanic",
      RIDRETH1 == 3 ~ "Non-Hispanic White",
      RIDRETH1 == 4 ~ "Non-Hispanic Black",
      RIDRETH1 == 5 ~ "Other Race"
    ),
    RIAGENDR = case_when(
      RIAGENDR == 1 ~ "Male",
      RIAGENDR == 2 ~ "Female"
    )
  )

# table with risk values and gender
table(demo_trigly$RIAGENDR, demo_trigly$RISK)

# table with risk values and race/ethnicity
table(demo_trigly$RIDRETH1, demo_trigly$RISK)

```

#### Visualizing heart disease risk by gender
```{r}
# filtering risk
demo_trigly_risk_gender <- demo_trigly %>%
  filter(RISK == 'high') %>% # setting fixed value, shiny app will use input select
  group_by(RIAGENDR) %>%
  summarize(count = n())

colors = c("#B80095", "#3008B1")

p <- plot_ly(demo_trigly_risk_gender, labels = ~RIAGENDR, values = ~count, type = 'pie',
             marker = list(colors = colors))
p <- p %>% layout(title = 'Heart disease risk by gender',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
p
```

#### Visualizing heart disease risk by gender and race
```{r}
# filtering risk
demo_trigly_gender_race <- demo_trigly %>%
  filter(RISK == 'high') # setting fixed value, shiny app will use input select

p <- ggplot(data = demo_trigly_gender_race, mapping = aes(x = RIDRETH1, fill = RIAGENDR)) +
  geom_bar(position = "dodge", mapping = aes(y = ..prop.., group = RIAGENDR)) +
  scale_fill_manual(values = c("#B80095", "#3008B1")) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, hjust = 1)
        ) +
  labs(title = "Heart disease risk by gender and race",
       subtitle = "Risk based on LDL level",
       x = "Race and Ethnicity",
       y = "Risk percentage %",
       fill = "Gender") 
  # transition_states(RIAGENDR) +
  #     ease_aes('linear')
  #   anim_save("../shiny/www/gganimate02.gif", animate(p))
ggplotly(p)
```

#### Coding Marital status
```{r}
demo_trigly <- demo_trigly %>%
  mutate(
    DMDMARTL = case_when(
      DMDMARTL == 1 ~ "Married",
      DMDMARTL == 2 ~ "Widowed",
      DMDMARTL == 3 ~ "Divorced",
      DMDMARTL == 4 ~ "Separated",
      DMDMARTL == 5 ~ "Never married",
      DMDMARTL == 6 ~ "Living with partner"
    )
  )

# table marital status and heart disease risk
table(demo_trigly$DMDMARTL,demo_trigly$RISK)

```

#### Visualizing heart disease risk by marital status
```{r}
# filtering risk
demo_trigly_marital <- demo_trigly %>%
  group_by(DMDMARTL,RISK) %>%
  summarize(count = n()) %>%
  drop_na()

p <- ggplot(data = demo_trigly_marital, mapping = aes(x = reorder(DMDMARTL,-count), y = count)) +
  geom_col(mapping = aes(fill = RISK)) +
  coord_flip() +
  scale_fill_manual(values = c("#FF3500", "#FF7D05", "#FFC300")) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 1)
        ) +
  labs(title = "Heart disease risk by marital status",
       subtitle = "Risk based on LDL level",
       x = "Marital status",
       y = "Total patients",
       fill = "Risk level")
  # transition_reveal(count) +
  #     ease_aes('linear')
  #   anim_save("../shiny/www/gganimate01.gif", animate(p))
ggplotly(p)
```

#### Education level and heart disease risk
```{r}
# table marital status and heart disease risk
table(demo_trigly$DMDHREDU,demo_trigly$RISK)

```

#### Visualizing Heart disease risk by education level
```{r}
demo_trigly_education <- demo_trigly %>%
  filter(DMDHREDU <= 5)

p <- ggplot(data = demo_trigly_education, mapping = aes(x = DMDHREDU, y = LBDLDL)) +
  geom_point(color = "#FFC300", alpha = 0.4) +
  geom_smooth(size = 1.1, method = "loess", se = FALSE) +
  scale_x_log10() +
  theme(strip.background = element_rect(
     fill="#FF7D05", size=1.5, linetype="solid"
     ),panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 1)
     )+
  facet_wrap(~RISK, ncol = 3) +
  labs(x = "Education Level (1 - lowest, 5 - highest)",
       y = "LDL Levels",
       title = "Heart disease risk by education level",
       subtitle = "Risk based on LDL level")
ggplotly(p)
```
#### Annual household income of patients in study
```{r}
table(demo_trigly$INDHHIN2)
```
#### Visualizing Heart disease risk by Annual household income
```{r}
p <- ggplot(data = demo_trigly, mapping = aes(x = INDHHIN2, fill = RISK)) +
  geom_density(alpha = 0.7) +
  scale_x_log10() +
  scale_fill_manual(values = c("#FF3500", "#FFC300", "#FF7D05")) +
  labs(x = "Annual Household Income (1 - lowest, 100 - highest)",
       y = "Risk score",
       title = "Heart disease risk by annual household income",
       subtitle = "Risk based on LDL level")
ggplotly(p)
```

#### Visualizing National Center for Health Statistics region
```{r}
# filtering Maryland region, where National Center for Health Statistics is located at
states <- map_data("state")

nchs_region <- states %>%
  filter(region == 'maryland')

p <- ggplot() +
  geom_polygon(data = states, aes(x=long,y=lat, group = group, fill=region),color="white", alpha = 0.3) +
  geom_polygon(data = nchs_region, aes(x=long,y=lat, group = group, fill=region),color="white") +
  labs(title = "National Center for Health Statistics - Maryland") +
  theme_void()
  guides(fill = FALSE)
ggplotly(p) %>% layout(showlegend = FALSE)
```
#### Visualizing National Center for Health Statistics region
```{r}
# National Center for Health Statistics coordenates 
m <- leaflet() %>%
  addTiles() %>%
  addMarkers(lng=-76.951720,
             lat=38.969450,
             label="National Center for Health Statistics",
             labelOptions = labelOptions(noHide=T))
m
```


